Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-06-11 paper.Rmd
minor edits

> Our theoretically driven model generates accurate forecasts, with base specification AUCs of 0.82 and 0.85 over one- and six-month windows, respectively, and AUCs as high as 0.92 in other specifications. Our model also consistently and sometimes dramatically outperforms the alternatives we test... Cederman and Weidmann (2017, 476) argue that "the hope that big data will somehow yield valid forecasts through theory-free 'brute force' is misplaced in the area of political violence." Our results lend some credence to this claim.

@@

Blair and Sambanis (2020) (B\&S hereafter) machine learning models, and specifically random forests, whose specifications are not theory-informed. For the moment, we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models. In short, they uphold their assumption that theory-guided empirical research produces better conflict predictions than machine learning inspired efforts that are necessarily ad hoc combinations of available variables. They arrive at this conclusion by examining the problem of predicting civil war onset. They report that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory including a \textit{kitchen sink} model with more than 1,000 covariates.

@@

\begin{itemize}
\item Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
\item Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
\item Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government on one side and opposition or rebel actors on the other. These are directed, thus making for four total covariates.
\item CAMEO: counts for all CAMEO event codes totaling $1,159$ covariates, which are mostly zero for any country in any month. 
\item Average: unweighted average of the predictions from the four models briefly described above.
\end{itemize}

@@

Aside from the visual impact of ROC smoothing, the underlying motivation and methods involved in smoothing are broader and involve parametric estimation of population ROC curves [@henley:1988; @henley:2014]. The main application is in the evaluation of medical tests, e.g. diagnostic imaging, with early methods dating back several decades and ongoing development of both parametric and non-parametric smoothing methods [e.g. @zou:etal:1997; @pulit:2016]; it is not established which methods should be preferable in a given application [@henley:2014]. Notably, the context in which ROC smoothing is used differs in important aspects from typical conflict research applications, e.g. our data is closer to a census of the population and with well-known dependencies like spatial and temporal correlation, rather than a random sample that is approximately i.i.d. It is thus not clear that smoothing is justified nor valid. We also note that smoothed ROC plots do
are not widely used outside of the narrow context mentioned above.

@@

B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. B&S, however, actually weights the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly, in the 6-month version of this file.] This appears to be an easily corrected coding error on the part of B&S.

@@

The "PITF Split Population" model appears to be incorrectly implemented, owing to a coding error. B&S describe it on page 20:

> The final approach is a random forest analog to split-population modeling. We first compute the average PITF predicted probability for each country across all years in our training set. We define those that fall in the bottom quartile as "low risk" and the rest as "high risk." We then run our escalation model on the high-risk and low-risk subsets separately, combining the results into a single random forest (column 4).

@@

What impact did the ROC smoothing have overall on the performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B&S Table 2)?  Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each of the gray points corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed}. The vertical bar in each plot marks the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible in average. The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the result B&S find, namely that the escalation model is generally superior.

@@

Both the Average and CAMEO models outperform the Escalation model in almost all instances. The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves and overturns when using traditional AUC-ROC calculations.

@@

4. B&S: "Overall, our results suggest that while measures of structural risk may improve predictive performance, the value they add is marginal and inconsistent. [...] Incorporating PITF thus significantly reduces or only slightly improves the performance of the escalation model, regardless of the approach we take. \ldots "

The most straightforward method of incorporating the annual structural PITF variables---adding them to the predictors of the Escalation RF model---strictly outperforms the Escalation Only model. Note that the two other combination models considered are both non-standard and that the "PITF Split Population" model does not incorporate structural information at all. We thus conclude that adding structural variables improves predictive performance.

@@

B&S in contrast fit very large forests with 100,000 trees in the basic model form, but where each tree only operates on a very small sub-sample (N=100 or 500), drawn without replacement, of the available training data. This approach only works due to the choice to use regression, not classification, trees. Trying to use classification trees with the other parameter settings not work at all because it is almost guaranteed that a sample of 100 from the ~11,000 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample. As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:

>>>>

> Our theoretically driven model generates accurate forecasts, with base specification AUCs of 0.82 and 0.85 over one- and six-month windows, respectively, and AUCs as high as 0.92 in other specifications. Our model also consistently and sometimes dramatically outperforms the alternatives we test. [...] Cederman and Weidmann (2017, 476) argue that "the hope that big data will somehow yield valid forecasts through theory-free 'brute force' is misplaced in the area of political violence." Our results lend some credence to this claim.

@@

B&S' analysis is based on the use of non-linear, non-deterministic machine learning models, and specifically random forests, one of which has a specification they argues is theory-based, and several others with more generic sets of covariates. Notably, the analysis is at the country-month or country-(6 month) levels and relies in large part on indicators derived from the ICEWS event data. For the moment, we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models. In short, they uphold their assumption that theory-guided empirical research produces better conflict predictions than machine learning inspired efforts that are necessarily ad hoc combinations of available variables. They arrive at this conclusion by examining the problem of predicting civil war onset. They report that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory, including a \textit{kitchen sink} model with more than 1,000 covariates.

@@

\begin{itemize}
\item Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model, for interactions between the government on one side and opposition or rebel actors on the other. 
\item Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation, for interactions between the government and opposition or rebels. These are directed, thus making for four directed dyads, which with four quad categories make for a total of sixteen covariates. 
\item Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for the same four directed dyads, for a toal of four covariates. 
\item CAMEO: counts for all CAMEO event codes over the four actor dyads, totaling $1,160$ covariates, which are mostly zero for any country in any month. 
\item Average: unweighted average of the predictions from the four models briefly described above.
\end{itemize}

@@

Aside from the visual impact of ROC smoothing, the underlying motivation and methods involved in smoothing are broader and involve parametric estimation of population ROC curves [@henley:1988; @henley:2014]. The main application is in the evaluation of medical tests, e.g. diagnostic imaging, with early methods dating back several decades and ongoing development of both parametric and non-parametric smoothing methods [e.g. @zou:etal:1997; @pulit:2016]; it is not established which methods should be preferable in a given application [@henley:2014]. Notably, the context in which ROC smoothing is used differs in important aspects from typical conflict research applications, e.g. our data is closer to a census of the population and with well-known dependencies like spatial and temporal correlation, rather than a random sample that is approximately i.i.d. It is thus not clear that smoothing is justified nor valid. We also note that smoothed ROC plots are not widely used outside of the narrow context mentioned above.

@@

B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly, in the 6-month version of this file.] This appears to be an easily corrected coding error on the part of B&S.

@@

The "PITF Split Population" model also appears to be incorrectly implemented, owing to a coding error. B&S describe it on page 20:

> The final approach is a random forest analog to split-population modeling. We first compute the average PITF predicted probability for each country across all years in our training set. We define those that fall in the bottom quartile as "low risk" and the rest as "high risk." We then run our escalation model on the high-risk and low-risk subsets separately, combining the results into a single random forest [...].

@@

What impact did the ROC smoothing have overall on the performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B&S Table 2)?  Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}). The vertical bar in each plot marks the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible in average. The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the result B&S find, namely that the escalation model is generally superior.

@@

Both the Average and CAMEO models outperform the Escalation model in almost all instances (see also Figure \ref{fig:table1-plot}). The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves and overturns when using conventional AUC-ROC calculations.

@@

4. B&S: "Overall, our results suggest that while measures of structural risk may improve predictive performance, the value they add is marginal and inconsistent. [...] Incorporating PITF thus significantly reduces or only slightly improves the performance of the escalation model, regardless of the approach we take. [...] "

The most straightforward method of incorporating the annual structural PITF variables---adding them to the predictors of the Escalation RF model---strictly outperforms the Escalation Only model. Note that the two other combination models considered are both non-standard and that the "PITF Split Population" model does not incorporate structural information at all, yet they also both do well. We thus conclude that adding structural variables improves predictive performance.

@@

B&S in contrast fit very large forests with 100,000 trees in the basic model form, but where each tree only operates on a very small sub-sample (N=100 or 500), drawn without replacement, of the available training data. This approach only works due to the choice to use regression, not classification, trees. Trying to use classification trees with the other parameter settings does not work at all because it is almost guaranteed that a sample of 100 from the ~11,000 training data rows with 9 positive cases will only include 0 (negative) outcomes in the sample. As it is, using regression with a 0 or 1 outcome produces warnings when estimating the models:
